---
title: "Cyclistic Data Analysis"
author: "Jeffrey Cuevas"
date: "1/3/2022"
output: pdf_document
---


### Overview
I'm a data analyst working on the marketing team at Cyclistic, a bike-share company in Chicago.  The director of marketing believes the company's future success depends on maximizing the number of annual memberships.  My team has been assigned the task of understanding how casual riders and annual members use Cyclistic bikes differently.  From these insights, our team will design a new marketing strategy to convert casual riders into annual members.

### Ask
Three questions will guide the future marketing program:

* How do annual members and casual riders use Cyclistic bikes differently?
* Why would casual riders buy Cyclistic annual memberships?
* How can Cyclistic use digital media to influence casual riders to become members?

This analysis will focus on how annual members and casual riders use Cyclistic bikes differently.

### Prepare
I used Cyclistic's historical trip data to analyze and identify trends.  Data used for this analysis consists of 12 individual .csv files representative of different months within the last year.  Data was merged and prepped using Pyspark within a Jupyter Notebook.  The output of the Jupyter Notebook will be used for analysis within R Studio.

### Process
List of operations used for processing the data:

1. Read each file into a dataframe and merge all files into one dataframe.
2. Removed duplicate rows (there were none).
3. Created new column called "distance_traveled" that uses Haversine formula to calculate distance between two points based on latitude and longitude.
4. Created new column called "date_diff" that calculates the day difference between ride start time and end time.
5. Removed records where "date_diff" < 0 because those are impossible scenarios.
6. Created new column called "duration_mins" that calculates the minute difference between ride start time and end time.
7. Removed records where "duration_mins" < 0 because those are impossible scenarios.
8. Created new column called "day_of_week" using pyspark date_format function.
9. Calculated frequency distribution for day of the week.
10. Calculated frequency distribution for casual and annual members.
11. Calculated frequency distribution for bike types.
12. Removed unnecessary columns and exported to a .csv file.

### Analyze

Let's take a look at weekly frequency distribution of the member and casual customers along with bike types:
```{r Libraries, include=FALSE}
library(tidyverse)
library(dplyr)
library(readr)
```
```{r member_casual_bike_types, echo=FALSE}
ds <- read.csv("ds_dropped.csv", header = T)
ds$day_of_week <- factor(ds$day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
ggplot(ds) + geom_bar(mapping=aes(x = day_of_week, fill = rideable_type))+
  facet_wrap(~member_casual)+
  theme(axis.text.x=element_text(angle=90, hjust=1))
```

* We can infer that members are primarily made up of working people since the distribution is pretty even throughout the week and lower on Sunday.
* Distribution for casual shows a pattern of high usage on the weekends and low usage during the week.
* Docked bikes have the highest usage among both groups.

Now, let's observe trip duration behavior for member and casual:
```{r member_casual_histogram, echo=FALSE}
ggplot(filter(ds, ds$duration_in_min < 100))+
  geom_histogram(mapping=aes(x = duration_in_min), bins = 30)+
  facet_wrap(~member_casual)
```

* Casuals take longer trips than members and members take shorter trips than casuals.
* The majority of both groups take shorter trips.

Lets plot distance traveled in meters for casuals and members:
```{r distance_traveled_density, echo=FALSE}
ggplot(filter(ds, ds$distance_traveled < 10000))+
  geom_density(mapping=aes(x = distance_traveled))+
  facet_wrap(~member_casual)
```

* It is difficult to make assumptions based on this plot.

We can observe a few summary statistics for both member and casual groups:

### Members
```{r member_summary, echo=FALSE, comment=NA}
#filter only members to a new dataset
filtered_member <- filter(ds, member_casual=="member")

#summary for member ds after dropping NA values
summary(drop_na(select(filtered_member, c('day_of_week', 'distance_traveled', 'duration_in_min'))))
```
### Casuals
```{r casual_summary, echo=FALSE, comment=NA}
#filter only casuals to a new dataset
filtered_casual <- filter(ds, member_casual=="casual")

#summary for casual ds after dropping NA values
summary(drop_na(select(filtered_casual, c('day_of_week', 'distance_traveled', 'duration_in_min'))))
```

* Mean distance traveled is relatively the same for both members and casuals.
* Median distance traveled is relatively the same for both members and casuals.
* Mean duration for casuals is almost three times more than members.
* Median duration for casuals is almost twice as much as members.

Next, lets find the most popular start and end stations with their frequency for members:

#### Most Popular Start Stations for Members
```{r member_start_station, echo=FALSE, comment=NA}
#sort the dataset by most popular start station name
head(count(filtered_member, start_station_name, sort=T), n=10)
```

#### Most Popular End Stations for Members
```{r member_end_station, echo=FALSE, comment=NA}
#sort the dataset by most popular end station name
head(count(filtered_member, end_station_name, sort=T), n=10)
```
* Members tend to start and end trips from the same stations.

#### Most Popular Start Stations for Casuals
```{r casual_start_station, echo=FALSE, comment=NA}
#sort the dataset by most popular start station name
head(count(filtered_casual, start_station_name, sort=T), n=10)
```

#### Most Popular End Stations for Casuals
```{r casual_end_station, echo=FALSE, comment=NA}
#sort the dataset by most popular end station name
head(count(filtered_casual, end_station_name, sort=T), n=10)
```
* Casuals tend to start and end trips from the same stations.

We can observe the most popular routes by looking at the most frequent combinations of start station and end station.
#### Most Popular Member Routes
```{r member_route, echo=FALSE, comment=NA}
#concat member start station name with end station name and observe the top 10 routes
filtered_member$routes <- paste(filtered_member$start_station_name, "-", filtered_member$end_station_name)
head(count(filtered_member, routes, sort=T), n=10)
```

#### Most Popular Casual Routes
```{r casual_route, echo=FALSE, comment=NA}
#concat casual start station name with end station name and observe the top 10 routes
filtered_casual$routes <- paste(filtered_casual$start_station_name, "-", filtered_casual$end_station_name)
head(count(filtered_casual, routes, sort=T), n=10)
```

#### Summary of All Findings
```{r summary_all, echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(formattable)
library(tidyr)
df <- data.frame('User_type' = c("Member", "Casual"),
                 "Amount" = c("2,352,923 (57.9%)", "1,710,107 (42.1%)"),
                 "Avg_and_median_trip_duration" = c("15.26 min - 11.07 min", "41.83 min - 20.20 min"),
                 "Avg_and_median_trip_distance" = c("2.25 km - 1.69 km", "2.19 km - 1.66 km"),
                 "Busiest_day" = c("Saturday", "Saturday"),
                 "Preferred_bike_type" = c("docked bike", "docked bike"),
                 "Most_occured_route" = c("Ellis Ave & 60th St - Ellis Ave & 55th St (1,409)",
                                          "Streeter Dr & Grand Ave - Streeter Dr & Grand Ave (8,230)"))
formattable(df,
            align = c('l', 'c', 'c', 'c', 'c', 'c', 'r'),
            list("User_type" = formatter("span", style = ~ style(color = "grey", font.weight = "bold"))))
```

### Share
The following are my main observations of this analysis:

* Members tend to take more rides during the week with a preference for docked bikes.  This leads me to believe that members are using Cyclistic bikes for commuting.
* Casuals tend to take more rides on the weekends with a preference for docked bikes.
* The majority of trips for both members and casuals is less than 25 minutes.  We can infer both user groups or using Cyclistic's bikes for short trips.
* Average distance traveled for both members and casuals is about the same, meaning there isn't much to draw from that.
* Average trip duration for casuals is almost three times higher than that of members.
* Both members and casuals appear to start and end trips from the same station.

### Act
How can we convert casuals to members?  There are a few ways to go about accomplishing this:

1. Offer weekend signup promotions for casual users.  This could be a limited time discount offer or a $0 signup fee.  Target the high traffic days (Friday - Sunday).
2. Provide an incentive program for casuals to signup for memberships.  This could come in the form of partnering with local businesses that would provide a free good or service upon signup.
3. Increase bike rental fees, especially for docked bikes, on the weekends to force casual members to consider buying a membership.
